---
title: "Project Proposal + Data"
author: 
  - "Jack Colt"
  - "Finn Balfe"
  - "John Brune"
  - "Alex Musco"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(janitor)
corruption <- read.csv("corruption_perception_index.csv")
happiness <- read.csv("hapiscore.csv")
```

## Data and Variable Description

To analyze the association between a country’s Corruption Perception Index (CPI) and World Happiness Report (WHR) we are using data obtained from Gapminder, a Swedish website that collects time series data from multiple sources to create unique datasets that cannot be found elsewhere. We are collecting data from Gapminder for both of our variables, with individual countries as the observational unit.

Our explanatory variable is the Corruption Perception Index, which is an index that is published annually by the non-government organization, Transparency International. The index ranks 180 countries on a scale of 0 (highly corrupt) to 100 (very clean) based on how corrupt their governments are each year. It is optimal for a country to have the highest score possible because that would indicate less corruption in their government.

Our response variable is a Happiness Score from the World Happiness Report, a partnership of Gallup, the Oxford Wellbeing Research Centre, the UN Sustainable Development Solutions Network, and the WHR’s Editorial Board, under which the report is published. Each observation is the national average response of life evaluations asking individuals to rate their current lives on a scale of 0 to 10 (with 10 meaning the individual has a perfect life) -- a question called the “Cantril Ladder.” Gapminder converted the observations to a scale of 0 to 100, which works well with our explanatory variable (CPI) as they are on the same scales in our datasets.

## Hypothesized Relationship Between Our Variables

We expect a positive relationship between Corruption Score and Happiness Score because we believe that countries with high corruption scores (meaning the country has low corruption) would experience more happiness overall. If a nation’s government is suspected to be corrupt, we hypothesize that it is more likely for their citizens to experience lower rates of happiness, since their government is ineffective. Corruption usually leads to a lower quality of life, since the economy is overall damaged and run inefficiently by the government’s misconduct, leading to lower potential output in the economic long run and lower living standards for its citizens. Corrupt economic operations, according to Elvin Mirzayev -- a writer for Investopedia, often “lead to monopolies or oligopolies in the economy.” Monopolies are objectively disadvantageous to buyers in the economy, and lead to high prices and lower quality products for consumers. Overall, we would expect a nation with high corruption (a low CPI) to have a low WHR score, while a nation with low corruption (high CPI) would experience higher WHR.

Citations: 
https://www.investopedia.com/articles/investing/012215/how-corruption-affects-emerging-economies.asp 

## Data Cleaning

```{r}
corruption <- corruption |>
  clean_names() |>
  pivot_longer(cols = c(x2012, x2013, x2014, x2015, x2016, x2017), 
               names_to = "year",
               values_to = "corruption") |>
  filter(!(is.na(corruption)))
```

*For our corruption data, we first used the janitor package to get our column names in tidy form. Next, we pivot longer the years from 2012 to 2017 because those are the only years we have corruption data for. Then, we remove any observations that contain missing values in the new corruption column. We chose to remove missing values after pivoting because, for countries that are only missing data for some of the years, we want to keep the data for the years we have data for.*

```{r}
happiness <- happiness |>
  clean_names() |>
  select(country, x2012, x2013, x2014, x2015, x2016, x2017) |>
  pivot_longer(cols = c(x2012, x2013, x2014, x2015, x2016, x2017), 
               names_to = "year",
               values_to = "happiness") |>
  filter(!(is.na(happiness)))
```

*For our happiness data, we first use the janitor package to get the column names in tidy form. Next, we select the country column and the years 2012 to 2017 because we only have corruption data for those years. Then, we pivot longer those years and remove observations that contain missing values in our new happiness columns for the same reason as the corruption data.*

```{r}
data <- happiness |>
  inner_join(corruption, join_by(country, year)) |>
  mutate(country = fct(country),
         year = fct(substr(year, 2, 5)))
```

*Finally, we inner join the happiness and corruption datasets on country name and year. We chose an inner join to remove country, year pairs that we do not have both happiness and corruption scores for. Lastly, we factor the country and year columns and remove the 'x' prefix from the year values.*

```{r}
glimpse(data)
```

